Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-13 paper/paper.qmd
Provide link to available data directly
Fix citation style again!
Updates according to SM's comments. (Not including connectivity related comments until confirmation!)

abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from thermal refugia that reduce bleaching impacts. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth refugia should be prioritized for further consideration in
    conservation planning to maximize coral survival under continuing climate change.

@@

The GBR is the world's largest coral reef ecosystem consisting of thousands of individual
reefs interconnected through larval dispersal patterns mediated by hydrodynamic, spatial,
and ecological processes [@cappo2000]. The GBR faces severe projected coral cover declines
under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover declines will be
compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs)
[@australianinstituteofmarinescience2024; @cantin2024].

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors that can either amplify or reduce thermal stress. Depth
is a crucial factor that can provide a refuge from thermal stress [@álvarez-noriega2018;
@baird2018; @smith2014]. Biological factors including coral size [@brandt2009], taxa
[@álvarez-noriega2018; @mcclanahan2005], and thermal tolerance through acclimation and
inheritance [@humanes2022; @lachs2024] all influence survival outcomes. Corals demonstrate
adaptive capacity to develop thermal tolerance that could delay widespread bleaching under
moderate emission scenarios [@bairos-novak2021; @humanes2022], representing a crucial
mechanism for reef persistence.

Beyond individual survival, reef-scale resilience depends on connectivity networks that
enable recolonization after disturbances [@angeler2023; @elmhirst2009; @hock2017]. Larval
source populations that export more larvae than they import are essential for system-wide
recovery, providing larvae that enable recolonization at distant locations [@hock2017;
@thomas2015]. Yet these source reefs are paradoxically more vulnerable individually, relying
heavily on local recruitment for their own recovery. When disturbances impact major source
populations, cascading effects reduce larval export and slow recovery throughout the
connected network [@figueiredo2022; @hock2017; @munday2009].

@@

Findings on the importance of connectivity likely differ from literature due to the highly
uncertain nature of larval connectivity modelling [@choukroun2025; @vogt-vincent2023], the
way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox, and the
magnitude of early bleaching events decreasing the breeding capacity of strongly connected
reefs. Analysing daily connectivity data Vogt-Vincent, Mitarai and Johnson
[-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for capturing
the seasonal - ecological time scale connectivity between locations due to high daily
variability in connectivity data. Connectivity matrices used in this investigation have
large relative standard deviations for larval transition probabilities between reefs over
the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that
capture ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three
GBR-scale connectivity models, including GBR1, and suggest the use of a model ensemble
approach to estimate larval dispersal due to the high sensitivity of connectivity
predictions to model specification. Further improvements and research in connectivity
modelling are necessary when incorporating connectivity as a supplementary factor in reef
resilience and intervention-planning analyses.

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program (<https://gbrrestoration.org>; RRAP -@thereef).

@@

The full reproducible investigation, and links to required data sources, can be found at
Grier and Iwanaga [-@grier2025].

>>>>

abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from reduced thermal exposure. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth should be considered as a key variable in
    conservation planning to maximize coral survival under continuing climate change.

@@

The GBR is the world's largest coral reef ecosystem consisting of thousands of individual
reefs interconnected through larval dispersal pathways mediated by hydrodynamic, spatial,
and ecological processes [@cappo2000]. The GBR faces severe projected coral cover declines
under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover declines will be
compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs),
representing the most widespread bleaching event on the GBR and largest single year decline
in the history of the Long Term Monitoring Program (LTMP;
@australianinstituteofmarinescience2024; @cantin2024).

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors that can either amplify or reduce the effects of
thermal stress. Depth is a crucial factor that can provide a refuge from thermal stress
[@álvarez-noriega2018; @baird2018; @smith2014]. Biological factors including coral size
[@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and thermal tolerance through
acclimation and inheritance [@humanes2022; @lachs2024] all influence survival outcomes.
Corals demonstrate adaptive capacity to develop thermal tolerance that could delay
widespread bleaching under moderate emission scenarios [@bairos-novak2021; @humanes2022],
representing a crucial mechanism for reef persistence.

Beyond individual survival, reef-scale resilience in part depends on connectivity networks
that enable recolonization after disturbances [@angeler2023; @elmhirst2009; @hock2017].
Larval source populations that export more larvae than they import are essential for
system-wide recovery, providing larvae that enable recolonization at distant locations
[@hock2017; @thomas2015]. Yet these source reefs are paradoxically more vulnerable
individually, relying heavily on local recruitment for their own recovery. When disturbances
impact major source populations, cascading effects reduce larval export and slow recovery
throughout the connected network [@figueiredo2022; @hock2017; @munday2009].

@@

Findings on the importance of connectivity likely differ from literature due to the highly
uncertain nature of larval connectivity modelling [@choukroun2025; @vogt-vincent2023] and
the way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox.
Additionally, and the high magnitude of early bleaching events (2025-2035) may decrease the
breeding capacity of strongly connected reefs, resulting in reduced importance of
connectivity in recovery. Analysing daily connectivity data Vogt-Vincent, Mitarai and
Johnson [-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for
capturing the seasonal - ecological time scale connectivity between locations due to high
daily variability in connectivity data. Connectivity matrices used in this investigation
have large relative standard deviations for larval transition probabilities between reefs
over the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that
capture ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three
GBR-scale connectivity models, including GBR1, and suggest the use of a model ensemble
approach to estimate larval dispersal due to the high sensitivity of connectivity
predictions to model specification. Further improvements and research in connectivity
modelling are necessary when incorporating connectivity as a supplementary factor in reef
resilience and intervention-planning analyses.

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program (<https://gbrrestoration.org>; RRAP [-@thereef]).

@@

The full reproducible investigation, and links to required data sources, can be found at
Grier and Iwanaga ([-@grier2025]; https://doi.org/10.5281/zenodo.17333564).
